---
title: "GIS_workshop_2025"
author: "Ronnie Bailey-Steinitz"
date: "2025-02-14"
output: html_document
editor_options: 
  chunk_output_type: console
---

# Introduction

Welcome to the workshop on **GPS Data Analysis & Visualization**! In this session, we will learn how to:

- Load and clean GPS data
- Visualize GPS points on maps
- Perform home range analysis (Minimum Convex Polygon and Kernel Density Estimation)
- Customize maps for reports
- Export results to use in QGIS

## Prerequisites
Before we start, make sure you have installed the following:

- **Software:** R, RStudio, QGIS
- **R Packages:**
  ```{r}
  install.packages(c("tidyverse", "sf", "sp", "rnaturalearth", "rnaturalearthdata", "ggspatial", "adehabitatHR"))
  ```

# 1. Loading and Cleaning GPS Data

First, we load our GPS dataset. The dataset should be a CSV file with columns for latitude, longitude, and individual ID.
(to add a code chunk, hit Cmd+Opt+i (Mac) / Ctrl+Alt+i (Windows & Linux))

```{r}
library(tidyverse)
library(sf)

gps_data <- read_csv("gps_data.csv")
head(gps_data)
```

we can also use publicly-available GPS datasets to practice: 
```{r}

install.packages("move")
library(move)

# Load an example dataset
data("leroy")
plot(leroy)

# Convert to a standard data frame
leroy_df <- as.data.frame(leroy) %>% 
  rename(longitude = location.long,
         latitude = location.lat)

leroy_df_filtered <- leroy_df %>%
  filter(longitude >= -180 & longitude <= 180,  # Valid longitudes
         latitude >= -90 & latitude <= 90)  # Valid latitudes


# View the first few rows
head(leroy_df)

# Show column names and data types
str(leroy_df)  

# place the Leroy dataset into out "gps_data" object to use for the exercise
gps_data <- leroy_df

```


Now, we convert the data into a **spatial object** so R recognizes it as geographic data.

```{r}
gps_sf <- st_as_sf(gps_data, coords = c("longitude", "latitude"), crs = 4326)
```

### Detecting Outliers
GPS errors can occur, so let’s check for outliers by plotting the raw data.

```{r}
library(ggplot2)

ggplot() +
  geom_sf(data = gps_sf, alpha = 0.6) + 
  theme_minimal()

# you don't need to tell ggplot what your lat/lon are (e.g., "aes(x = lon, y = lat)") because we converted a spreadsheet into an 'sf' object with "geometry", which is inherently geographic- it already has embedded coordinates.

# have a look: 
class(gps_sf$geometry)
# [1] "sfc_POINT" "sfc"     

```

If you see points far from expected locations, you may need to filter extreme values.


# 2. Visualizing GPS Data on a Map

We will use **Natural Earth Data** (an open-access map source) to plot GPS points.

```{r}
# install.packages("rnaturalearthdata")
library(rnaturalearth)
library(ggspatial)

world <- ne_countries(scale = "medium", returnclass = "sf")

# Compute bounding box and expand it
bbox <- st_bbox(gps_sf)

# Expand the bbox slightly (adjust factor as needed)
expand_factor <- 100  # Expands by 5%
expand_factor <- 0.05  # Expands by 100%
bbox_expanded <- bbox + c(-expand_factor, -expand_factor, expand_factor, expand_factor) * (bbox["xmax"] - bbox["xmin"])

# Convert bbox to an sf polygon for plotting
bbox_sf <- st_as_sfc(st_bbox(bbox_expanded, crs = st_crs(gps_sf)))

ggplot() +
  geom_sf(data = world, fill = "gray95", color = "gray70") +
  geom_sf(data = gps_sf, alpha = 0.6, size = 2) +
  coord_sf(xlim = c(bbox_expanded["xmin"], bbox_expanded["xmax"]), 
           ylim = c(bbox_expanded["ymin"], bbox_expanded["ymax"])) +
  theme_minimal() +
  annotation_scale(location = "br") +
  annotation_north_arrow(location = "tl", which_north = "true")


```


Using ggmap for satellite basemap

```{r}

install.packages("basemaps")
library(basemaps)

# check out what maps they have
get_maptypes()

# Get bounding box of gps_data
bbox <- st_bbox(gps_sf)

ggplot() +
  basemap(bbox, basemap = "natgeo_world_map") +
  geom_sf(data = gps_sf, fill = NA) +
  coord_sf(xlim = c(bbox$xmin, bbox$xmax), ylim = c(bbox$ymin, bbox$ymax), crs = 4326)



ggplot() +
  bm_map() +  # Adds the Esri NatGeo World Map as a basemap
  geom_sf(data = gps_data, color = "blue", alpha = 0.6) +  # GPS points
  geom_sf(data = st_as_sfc(bbox_expanded), fill = NA, color = "red", linetype = "dashed") +  # Bounding box
  coord_sf(xlim = c(bbox_expanded["xmin"], bbox_expanded["xmax"]), 
           ylim = c(bbox_expanded["ymin"], bbox_expanded["ymax"])) +
  theme_minimal()






```


### Customizing the Visualization
- **Change colors by individual**
- **Adjust point size and opacity**

# 3. Home Range Analysis

### 3.1 Minimum Convex Polygon (MCP)
MCP is a simple method that creates a convex shape around an animal’s locations to estimate its home range.

```{r}
library(adehabitatHR)

coords <- gps_sf %>% select(geometry) %>% st_coordinates()
mcp_poly <- mcp(as.ltraj(coords, id = gps_sf$individual_id, typeII = FALSE), percent = 95)

ggplot() +
  geom_sf(data = world, fill = "gray95", color = "gray70") +
  geom_sf(data = mcp_poly, fill = "blue", alpha = 0.3, color = "blue") +
  geom_point(data = gps_sf, aes(x = longitude, y = latitude), color = "red", alpha = 0.6) +
  theme_minimal()
```

### 3.2 Kernel Density Estimation (KDE)
KDE estimates an animal’s home range by calculating the probability of where it spends time. Unlike MCP, KDE accounts for varying densities of points.

```{r}
kde_res <- kernelUD(coords, h = "href", grid = 500)
kde_poly <- getverticeshr(kde_res, 95)

ggplot() +
  geom_sf(data = world, fill = "gray95", color = "gray70") +
  geom_sf(data = kde_poly, fill = "purple", alpha = 0.3, color = "purple") +
  geom_point(data = gps_sf, aes(x = longitude, y = latitude), color = "red", alpha = 0.6) +
  theme_minimal()
```

# 4. Importing Results into QGIS
- Export gps_data as .csv
- Export **MCP and KDE home ranges** as shapefiles
- Load shapefiles into QGIS
- Customize colors, transparency, and labels

# Exporting the dataset as .csv
```{r}

write.csv(gps_data, "leroy_gps_data.csv")

```


# 5. Additional Visualizations

### **Tracking Animal Movements Over Time**
```{r}
ggplot() +
  geom_sf(data = world, fill = "gray95", color = "gray70") +
  geom_path(data = gps_sf, aes(x = longitude, y = latitude, group = individual_id, color = individual_id)) +
  theme_minimal()
```

### **Heatmap of GPS Density**
```{r}
# Example of density estimation visualization
library(ggplot2)
ggplot(gps_sf, aes(x = longitude, y = latitude)) +
  stat_density_2d(aes(fill = ..level..), geom = "polygon") +
  theme_minimal()
```

# Conclusion
This workshop introduced the basics of GPS data analysis using R and QGIS. You can explore more by testing different home range techniques, adding environmental layers in QGIS, or using time-based movement analysis.

**Next Steps:**
- Try mapping other species or datasets
- Explore different projection systems in QGIS
- Combine GPS data with habitat or environmental layers

Happy coding! 🚀